---
title: "Introduction to RLSPrivate"
author: "CJ Brown"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results ='markup', warnings = FALSE, comment = '', strip.white = FALSE)
```  

This package provides access to the [Reef Life Survey](https://reeflifesurvey.com/) data-base and helpful functions for analyising and visualising this extensive data source.

The citation for this data is:
Edgar GJ, Stuart-Smith RD. Systematic global assessment of reef fish communities by the Reef Life Survey program. Scientific Data. 2014 May 27;1:140007.

## Getting started

Once you have installed `RLSPrivate` then you can get access to the databases:

```{r}
    library(RLSPrivate)
    data(rls)
    data(sdat)
    data(survdat)
    data(fdat)
```
`rls` is the main database of observations, `sdat` is a database of site covariates and `survdat` is a database of survey level covariates. 

Weights are by default recorded in grams, so you may want to convert them to kg first:
```{r}
rls$BioMass <- rls$BioMass/1000 #convert to kg
```


## All metrics

The general purpose function `allmetrics` will calculate a range of metrics of different types (biomass, diversity and functional groups) for a given grouping variable. Below we calculate all the metrics by `SurveyID`.  We pass in two internal dataframes `cryptdat` and `invertclasses` which just specify which families belong to cryptic fish and invertebrates. 

```{r}
data(cryptdat)
data(invertclasses)

dout <- allmetrics(rls, SurveyID, sdat = sdat, survdat = survdat, fdat = fdat, cryptdat = cryptdat, invertclasses = invertclasses)
dout
```

## Diversity and biomass metrics 

You can calculate a range of metrics using the `calcmetrics` function, such as the alpha diversity: 

```{r}
   calcmetrics(rls, metrics = c("alpha"))
```  

This will pass a warning to let you know you are calculating alpha diversity across multiple methods. You could filter `rls` for a single method, see example for functional groups below. 

The data-frame `calcmetrics` can also take a grouping variable as input (must be a named variable in the data.frame at hand). For instance, to calculate the biomass of fish >40cm: 

```{r}
 head(calcmetrics(rls, SurveyID, metrics = c("bioatsize"), sizecat = 40))
```

You could also join `survdat` to site level data to calculate site level alpha diversity (i.e. average number of fish observed on all surveys at that site): 

```{r}
    #alpha diversity
    rls %>%
        inner_join(survdat) %>%
        calcmetrics(SiteCode, metrics = "alpha")
```

If you want to know gamma diversity (total number of species observed at a site), then use the `nspp` function:  

```{r}
    #gamma diversity by sites 
    rls %>%
        inner_join(survdat) %>%
        nspp(SiteCode)

    #gamma diversity by ecoregions
   rls %>%
        inner_join(survdat) %>%
        inner_join(sdat) %>%
        nspp(Ecoregion)

```

## Functional group metrics 

If you want to get biomass of different functional groups, then use `tgroup_metrics()`:  

```{r}
flim <- fdat %>% select(SpeciesID, Class, TrophicGroup)

herbs <- c("browsing herbivore", "scraping herbivore", "excavator", "algal farmer")

r2 <- rls %>% filter(Method ==1) %>%
    inner_join(survdat) %>%
    inner_join(sdat)
fout <- tgroup_metrics(r2, Location, Ecoregion, tgroups = herbs, traits = flim, na.rm = T, metrics = "biomass")

hist(fout$biomass)
```

## Community temperature index

The community temperature index is calculated for a survey as:
$$\frac{\sum^n_i{T_j B_{i,j}}}{\sum^n_i{B_{i,j}}}$$
Where there are $n$ surveys $i$,$T_j$ are the species thermal indexes (e.g. thermal midpoint from a species range) and $B_{i,j}$ are the biomasses or abundances of each species at each site.

For instance, to calculate the CTI for locations and ecoregions: 

```{r}
flim <- fdat %>% select(SpeciesID, Class, Family, ThermalMP_5_95)

r2 <- rls %>% filter(Method ==1) %>%
    inner_join(survdat) %>%
   inner_join(sdat) %>%
    inner_join(flim) %>%
    filter(Class %in% fish_classes())
 xout <- r2 %>% commtempindex(Location, Ecoregion, tempvar = "ThermalMP_5_95", abundvar = "BioMass", dolog = TRUE)
hist(xout$CTI)

```
